---
title: "Stroke - exploratory data analysis"
output: html_notebook
---

```{r}
# Load libraries
library(skimr)
library(here)
library(tidyverse)
```

```{r}
# Read in stroke hospital data
activity_hb <- read_csv(here("data/clean/activity_hb.csv")) %>% 
  filter(admission_type == "All")
activity_ca <- read_csv(here("data/clean/activity_ca.csv")) %>% 
    filter(admission_type == "All")
```

```{r}
activity_hb %>% 
  distinct(admission_type)

activity_hb %>% 
  distinct(age_group)

activity_hb %>% 
  distinct(diagnosis)

```
Possible sources of analysis:
- by diagnosis
- by age group
Unlikely to use by admission type.

```{r}
activity_hb %>% 
  filter(hb_name == "Scotland" & 
           admission_type == "All" &
           age_group =="All" & 
           sex == "All" & 
           year == "2018" ) %>% 
  group_by(diagnosis, sex, age_group) %>% 
  summarise(discharges = sum(number_of_discharges))

```

```{r}
activity_hb %>% 
  filter(hb_name == "Scotland"  & 
           admission_type == "All" &
           age_group =="All" & 
           year == "2018") %>% 
  group_by(diagnosis, sex, age_group) %>% 
  summarise(discharges = sum(number_of_discharges))

```

```{r}
activity_hb %>% 
  filter(hb_name == "Scotland"  & 
           admission_type == "All" & 
           year == "2018") %>% 
  group_by(diagnosis, sex, age_group) %>% 
  summarise(discharges = sum(number_of_discharges))
```

Initial plots to see shape of data:

```{r}
activity_hb %>% 
  skim()
```

```{r}
mortality_ca <- read_csv(here("data/clean/mortality_ca.csv"))
skim(mortality_ca)
```

```{r}
mortality_hb <- read_csv(here("data/clean/mortality_hb.csv")) 

mortality_hb %>% 
  filter(is.na(number_of_deaths)) %>% 
  distinct(hb_name)
```


```{r}
activity_hb %>% 
  filter(hbr == "S92000003" &
           admission_type == "All" &
           age_group == "All" &
           sex == "All"
         ) %>% 
  ggplot(aes(x = year, y = number_of_discharges, group = diagnosis)) +
  geom_line()
```

```{r}
mortality_hb %>% 
  filter(hbr == "S92000003" &
           age_group == "All" &
           sex == "All"
         ) %>% 
  ggplot(aes(x = year, y = number_of_deaths, group = diagnosis)) +
  geom_line()
```

```{r}
activity_hb %>% 
  filter(hb_name == "Scotland" &
           admission_type == "All" &
           age_group == "All" &
           sex != "All") %>% 
  group_by(year, sex, diagnosis) %>% 
  summarise(total_cases = sum(number_of_discharges))
```
 
```{r}
mortality_hb %>% 
  filter(hb_name == "Scotland" & age_group == "All" & sex != "All") %>% 
  group_by(year, sex, diagnosis) %>% 
  summarise(total_deaths = sum(number_of_deaths))
```
 
 
 #-------------------------------
 High-level summary of data - discharges
  - look at totals for Scotland for 2018
  - first: by type of diagnosis (include CVD)
  - exclude CVD: 
    -show by sex
    -show by age band
 
 
```{r}
# Total discharges split by diagnosis, 2018
activity_hb %>% 
  filter(hb_name == "Scotland"
           & age_group =="All" 
           & sex == "All" 
           & year == "2018"
         ) %>% 
  group_by(diagnosis
           ) %>% 
  summarise(discharges = sum(number_of_discharges)) %>% 
  ggplot(aes(x = reorder(diagnosis, discharges),
             y = discharges
             )
         ) +
  geom_col() +
  coord_flip()
```
 
```{r}
# Total discharges by sex, diagnosis
activity_hb %>% 
  filter(hb_name == "Scotland"
           & age_group =="All" 
           & sex != "All" 
           & year == "2018"
         ) %>% 
  group_by(diagnosis, 
           sex
           ) %>% 
  summarise(discharges = sum(number_of_discharges)) %>% 
  ggplot(aes(x = reorder(diagnosis, discharges),
             y = discharges,
             fill = sex
             )
         ) +
  geom_col(position = "dodge") +
  coord_flip()
```
 
```{r}
# Total discharges by sex, diagnosis, age_group
activity_hb %>% 
  filter(hb_name == "Scotland"
           & !age_group %in% c("All", "under75 years") 
           & sex != "All"
           & year == "2018"
         ) %>% 
  group_by(diagnosis, 
           sex,
           age_group
           ) %>% 
  summarise(discharges = sum(number_of_discharges)) %>% 
  ggplot(aes(x = age_group,
             y = discharges,
             fill = sex
             )
         ) +
  geom_col(position = "dodge") +
  coord_flip() +
  facet_wrap(~ diagnosis)
    
  
```
 
  #-------------------------------
 High-level summary of data - deaths
  - look at totals for Scotland for 2018
  - first: by type of diagnosis (include CVD)
  - exclude CVD: 
    -show by sex
    -show by age band
    
    
```{r}
# Mortality by diagnosis
mortality_hb %>% 
  filter(hb_name == "Scotland"
           & age_group =="All" 
           & sex == "All" 
           & year == "2018"
         ) %>% 
  group_by(diagnosis
           ) %>% 
  summarise(deaths = sum(number_of_deaths)) %>% 
  ggplot(aes(x = reorder(diagnosis, deaths),
             y = deaths
             )
         ) +
  geom_col() 
```

```{r}
mortality_hb %>% 
  filter(hb_name == "Scotland"
           & age_group =="All" 
           & sex != "All" 
           & year == "2018"
         ) %>% 
  group_by(diagnosis,
           sex
           ) %>% 
  summarise(deaths = sum(number_of_deaths)) %>% 
  ggplot(aes(x = reorder(diagnosis, deaths),
             y = deaths,
             fill = sex
             )
         ) +
  geom_col(position = "dodge") 
```

```{r}
mortality_hb %>% 
  filter(hb_name == "Scotland"
           & !age_group  %in% c("All", "under75 years")
           & sex != "All" 
           & year == "2018"
         ) %>% 
  group_by(diagnosis,
           sex,
           age_group
           ) %>% 
  summarise(deaths = sum(number_of_deaths)) %>% 
  ggplot(aes(x = age_group,
             y = deaths,
             fill = sex
             )
         ) +
  geom_col(position = "dodge") +
 
  facet_wrap(~ diagnosis)
```

# Analysis of mortality rates for Scotland
- Overall mortality rates
- by sex
- by age and sex

```{r}
# Mortality rates
mortality_hb %>% 
  filter(hb_name == "Scotland"
           & !age_group  %in% c("All", "under75 years") 
           & sex != "All" 
           & year == "2018"
         ) %>% 
  ggplot(aes(x = age_group,
             y = easr,
             colour = sex)) +
  geom_point() +
  facet_wrap(~ diagnosis, scales = "free_y")
```

```{r}
# Health Board comparison

mortality_hb %>% 
  filter(age_group == "All"
           & sex == "All" 
           & year == "2018"
           & diagnosis == "Stroke"
         ) %>% 
  ggplot(aes(x = reorder(hb_name, easr),
             y = easr,
             colour = hb_name)) +
  coord_flip() +
  geom_point() 
```

```{r}
# Health Board comparison - stroke only
mortality_hb %>% 
  filter(diagnosis == "Stroke"
           & age_group == "75plus years" 
           & sex == "All" 
           & year == "2018"
         ) %>% 
  ggplot(aes(x = reorder(hb_name, easr),
             y = easr,
             colour = hb_name)) +
  geom_point() +
  coord_flip()
```


```{r}
mortality_ca %>% 
    filter(diagnosis == "Stroke"
           & age_group == "75plus years" 
           & sex == "All" 
           & year == "2018"
           & hb_name == "NHS Greater Glasgow and Clyde"
         ) %>% 
  ggplot(aes(x = reorder(ca_name, easr),
             y = easr,
             colour = ca_name)) +
  geom_point() +
  coord_flip() 
```

```{r}
mortality_ca %>% 
    filter(diagnosis == "Stroke"
           & age_group == "75plus years" 
           & sex == "All" 
           & year == "2018"
         ) %>% 
  ggplot(aes(x = reorder(hb_name, easr),
             y = easr,
             colour = ca_name)) +
  geom_point() +
  coord_flip() 
```

